A demonstration of the \LaTeXe\ class file for \itshape{SAGE Publications}

### 2015-12-04 body/appendix_algo.tex
corrections andrea

At each step, $g$ is called with the previous state as a parameter, as well
as a new root placement. $g$ returns a new contact configuration, if it suceeded
in computing a balance configuration with only one contact switch occuring.
If $g$ failed in achieving this, the method \textsc{RepositionContacts} is called.
It repositions one end effector (either a free limb, or the oldest active contact) towards a new balanced contact position if possible.
This repositionning allows to increase the odds that the contact can be maintained at the next step.

@@

\begin{algorithm}
\caption{Full body contact generation method} \label{interpolate}
	\begin{algorithmic}[1]
	%~ \Function{GenerateConfiguration}{}
	\Function{$g$}{$pState$,$\mathbf{q}^0$}
		\State State $newState$
		\State $newState.q0 = \mathbf{q}^0$
		\State $newState.freeLimbs = pState.freeLimbs$
		\State /*First try to maintain previous contacts*/
		\State $nbContactsBroken = 0$
		\For {\textbf{each} Limb $k$ in $previous.contactLimbs$}
			\If {$!$\textsc{MaintainContact}$(pState,\mathbf{q}^0,k)$}
				\State $nbContactsBroken += 1$
				\If {$nbContactsBroken > 1$}				
					\State \textbf{return} $NULL$
				\EndIf				
				\State $push(newState.freeLimbs,k)$
			\Else 					
				\State $push(newState.contactLimbs,k)$
			\EndIf
		\EndFor
		\For {\textbf{each} Limb $k$ in $previous.freeLimbs$}
			\State /*Generate a balanced contact*/
			\If {\textsc{GenerateContact}$(\mathbf{q}^0,k)$}	
				\State $push(newState.contactLimbs,k)$			
				\State \textbf{return} $newState$
			\EndIf
		\EndFor
		\State \textbf{return} $NULL$
	\EndFunction
\end{algorithmic}
\label{alg:pi}
\end{algorithm}

@@

\textsc{GenerateContact}$(\mathbf{q}^0,k)$ is a call to the contact generator presented in Section~\ref{sec:single_contact}. It generates a balanced contact configuration, and assigns the corresponding configuration to $k$.
If it fails, $k$ remains unchanged if it is collision free, otherwise it is assigned a random collision free configuration.

>>>>

At each step, $g$ is called with the previous state as a parameter, as well
as a new root placement. $g$ returns a new contact configuration, if it suceeded
in computing a configuration with only one contact switch occuring.
If $g$ failed in achieving this, the method \textsc{RepositionContacts} is called.
It repositions one end effector (either a free limb, or the oldest active contact) towards a new contact position if possible.
This repositionning allows to increase the odds that the contact can be maintained at the next step.

@@

\begin{algorithm}
\caption{Full body contact generation method} \label{interpolate}
	\begin{algorithmic}[1]
	%~ \Function{GenerateConfiguration}{}
	\Function{$g$}{$pState$,$\mathbf{q}^0$}
		\State State $newState$
		\State $newState.q0 = \mathbf{q}^0$
		\State $newState.freeLimbs = pState.freeLimbs$
		\State /*First try to maintain previous contacts*/
		\State $nbContactsBroken = 0$
		\For {\textbf{each} Limb $k$ in $previous.contactLimbs$}
			\If {$!$\textsc{MaintainContact}$(pState,\mathbf{q}^0,k)$}
				\State $nbContactsBroken += 1$
				\If {$nbContactsBroken > 1$}				
					\State \textbf{return} $NULL$
				\EndIf				
				\State $push(newState.freeLimbs,k)$
			\Else 					
				\State $push(newState.contactLimbs,k)$
			\EndIf
		\EndFor
		\For {\textbf{each} Limb $k$ in $previous.freeLimbs$}
			\If {\textsc{GenerateContact}$(\mathbf{q}^0,k)$}	
				\State $push(newState.contactLimbs,k)$			
				\State \textbf{return} $newState$
			\EndIf
		\EndFor
		\State \textbf{return} $NULL$
	\EndFunction
\end{algorithmic}
\label{alg:pi}
\end{algorithm}

@@

\textsc{GenerateContact}$(\mathbf{q}^0,k)$ is a call to the contact generator presented in Section~\ref{sec:single_contact}. It generates a contact configuration in static equilibrium, and assigns the corresponding configuration to $k$.
If it fails, $k$ remains unchanged if it is collision free, otherwise it is assigned a random collision free configuration.

### 2015-12-04 body/contact.tex
corrections andrea

In this Section we first describe a single contact generation process, that is how to generate a contact configuration for a given limb 
and root location.
Then, on this method we build an iterative algorithm to generate a discrete sequence of balanced contact configurations along a truly feasible path.

In what follows a \textit{valid} contact configuration is understood as being collision free, respecting joint limits, and allowing to maintain the quasi static balance
of the robot. The first two tests are trivial; our criterion to assert efficiently the quasi static balance of the system
is described in Section TODO.

@@

At each step, we consider a new root placement $\mathbf{q}^0_{i}$ and try to generate a balanced contact configuration that preserves as much as possible
the previous contacts.

@@

Figure~\ref{fig:break_contact} illustrates the contact persistence method.
If possible, a limb in contact at time $i-1$ remains in contact at $i$. The contact is broken if an inverse kinematics solver fails to find a collision free limb configuration that satisfies joint limits. The solver is directy provided by the Hpp software.

@@

%\subsection{Generation of a contact configuration}  
\subsection{Creating contacts.}
Contacts are created using three rules:
\begin{enumerate}
\item Only one contact creation can happen between two consecutive states. This increases the odds that the interpolation between the two states will be feasible; 
\item A contact is validated if and only if the resulting configuration is statically balanced, no matter what user-defined heuristics are used; 
\item We use a FIFO approach:  we always try to first create a contact with the limb that has been free for the longest time. If the contact creation
was not successful for a limb, the limb is pushed on top of the queue, and will only be tried again after the others have been treated.
\end{enumerate}

These three rules provide a deterministic manner to generate contacts along the discretized guide path.
FIGURE. For a given truly feasible path, our current implementation does not guarantee that the planner will succeed in generating a balanced configuration, because
of the deterministic approach used to break the combinatorial.
However, in practice the planner is successful in the large majority of cases, as discussed in section~\ref{sec:stat1}, thanks to 
the heuristics we use for relevant contact selection.

>>>>

In this Section we first describe a single contact generation process, that is how to generate a contact configuration for a given limb 
and root location.
Then, on this method we build an iterative algorithm to generate a discrete sequence of contact configurations in equilibrium, along a truly feasible path.

In what follows a \textit{valid} contact configuration is understood as being collision free, respecting joint limits, and allowing to maintain the static equilibrium
of the robot. The first two tests are trivial; our criterion to assert efficiently the static equilibrium of the system
is described in Section TODO.

@@

At each step, we consider a new root placement $\mathbf{q}^0_{i}$ and try to generate a contact configuration that preserves as much as possible
the previous contacts, and is in static equilibrium.

@@

Figure~\ref{fig:break_contact} illustrates the contact persistence strategy.
If possible, a limb in contact at time $i-1$ remains in contact at $i$. The contact is broken if an inverse kinematics solver fails to find a collision free limb configuration that satisfies joint limits. The solver is directy provided by the Hpp software.

@@

%\subsection{Generation of a contact configuration}  
\subsection{Creating contacts.}
Contacts are created using three rules:
\begin{enumerate}
\item Only one contact creation can happen between two consecutive states. This increases the odds that the interpolation between the two states will be feasible; 
\item A contact is validated if and only if the resulting configuration is in static equilibrium, no matter what user-defined heuristics are used; 
\item We use a FIFO approach:  we always try to first create a contact with the limb that has been free for the longest time. If the contact creation
was not successful for a limb, the limb is pushed on top of the queue, and will only be tried again after the others have been treated.
\end{enumerate}

These three rules provide a deterministic manner to generate contacts along the discretized guide path.
FIGURE. For a given truly feasible path, our current implementation does not guarantee that the planner will succeed in generating a configuration in static equilibrium, because
of the deterministic approach used to break the combinatorial.
However, in practice the planner is successful in the large majority of cases, as discussed in section~\ref{sec:stat1}, thanks to 
the heuristics we use for relevant contact selection.

### 2015-12-04 body/equilibrium.tex
corrections andrea

\subsection{A heuristic for robust quasi static balance}
The planner is designed so that any contact configuration it outputs is quasi statically balanced.
Balance is of course critical in multipedal locomotion. For this reason we are interested in a robust 
criterion, that ensures that the robot remains balanced in a real world application, regardless of the errors
that can result from peception and control uncertainties.

We first give a classic linear program (LP) that asserts that a contact configuration is statically balanced.
From this formulation we derive a new LP that returns a quantification of the robustness of the configuration regarding balance.
From this value we can in turn either define a heuristic to choose the more robust candidate, or set a treshold on the required balance robustness
for any configuration. While the LP we propose is new, it is based on an analysis of the problem we propose in \cite{Prete2016}, where the interested reader
can find more details.

@@

According to Coulomb's law, the robot is statically balanced if all contact forces lie in the friction cone defined by the surface.
As classically done, we linearize the friction cone in a conservative fashion with a pyramid, described by four generating rays. We choose for instance:
\begin{equation*}
\mathbf{V}_{i} = \mat{\mathbf{n}_{i} + \mu \mathbf{t}_{i1} & \mathbf{n}_{i} -\mu \mathbf{t}_{i1} & \mathbf{n}_{i} + \mu \mathbf{t}_{i2} & \mathbf{n}_{i} - \mu \mathbf{t}_{i2}}^T
\end{equation*}

Any force belonging to the linearized cone
can thus be express as a linear combination of its four generating rays.
Therefore a constraint on the $\mathbf{f}_i$ is given by the relation:

\begin{equation*}
\forall i  \qquad  \exists \bm{\beta}_i \in \mathbb{R}^{+4} , \mathbf{f}_{i} = \mathbf{V}_{i} \bm{\beta}_i \qquad 
\end{equation*}

we can stack the constraints to obtain:
\begin{equation}\label{eq:gen}
\exists \bm{\beta} \in \mathbb{R}^{+4e} , \mathbf{f} = \mathbf{V} \bm{\beta} \qquad 
\end{equation}
where $\mathbf{V} = \diag{ \{\mathbf{V}_1, \dots, \mathbf{V}_e\} }$, and $\mathbf{f} = [\mathbf{f}_0,...,\mathbf{f}_e]^T$.

Additional constraints on the forces result from Newton-Euler equations:

\begin{align}
\underbrace{\mat{m (\ddot{\mathbf{c}} - \mathbf{g})  \\ m \mathbf{c} \times (\ddot{\mathbf{c}} - \mathbf{g}) + \dot{\mathbf{L}}}}_\mathbf{w}
= 
\underbrace{
\mat{\mathbf{I}_3 & \dots & \mathbf{I}_3 \\
\hat{\mathbf{p}}_1 & \dots & \hat{\mathbf{p}}_e} \mathbf{V}
}_\mathbf{G}
\bm{\beta},
\end{align}
where $\mathbf{w}\in \Rv{6}$ is the so-called gravito-inertial wrench (GIW) \citep{qiu:dhm:2011, Caron2015} and $\hat{\mathbf{p}} \in \R{3}{3}$ is the cross-product matrix associated to $\mathbf{p}$.

Static balance assumes zero acceleration thus we can write:
\begin{align} \label{eq:new_eul}
\mathbf{w}_0 = 
\underbrace{\mat{\mathbf{0}_{3\times 3} \\ m \hat{\mathbf{g}}}}_{\mathbf{D}} \mathbf{c} + 
\underbrace{\mat{-m\mathbf{g} \\ \mathbf{0}}}_{\mathbf{d}}
\end{align}
The wrench $\mathbf{w}_0$ is independent of $c^z$ because of the cross product matrix $\hat{\mathbf{g}}$ inside $\mathbf{D}$, so we can rewrite \eqref{eq:new_eul} as:
\begin{align} \label{eq:new_eul2d}
\mathbf{w}_0 = \mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d}
\end{align}

@@

\begin{equation} \label{eq:lin_prog} \begin{aligned}
\find \quad & \bm{\beta} \in \Rv{+4e} \\
\st \quad &\mathbf{G} \bm{\beta} = \mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d} \\
& \bm{\beta} \ge 0 \\
\end{aligned} \end{equation}

\subsubsection{Formulation of a robust LP:}
Let $b_0 \in \mathbb{R}$ a scalar value. We now define the following LP:

\begin{equation} \label{eq:lin_prog_rob} \begin{aligned}
\find \quad & \bm{\beta} \in \Rv{+4e}, b_0 \in \Rv{} \\
\minimize  \quad & -b_0 \\
\st \quad &\mathbf{G} \bm{\beta} = \mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d} \\
& \bm{\beta} \ge b_0 \\
\end{aligned} \end{equation}

We observe that if $b_0$ is positive, LP~\ref{eq:lin_prog} admits a solution. The higher $b_0$, the further away from the cone limits admissible forces can be found. If $b_0$ is negative, the configuration is not balanced, and the value $b_0$ indicates "how far" from static balance
the configuration is. $b_0$ can thus be seen as a measure of robustness.

In our implementation, we solve the dual of LP~\ref{eq:lin_prog_rob}.
after a trivial elimination of Lagrange multipliers the dual problem has lower dimensionality:
\begin{equation} \label{eq:dual} \begin{aligned}
\find \quad & \bm{\nu} \in \Rv{6}\\
\minimize  \quad & (\mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d})^T \nu \\
\st \quad &\mathbf{G}^T \bm{\nu} \ge 0 \\
& \mathbf{1}^T \mathbf{G}^T \bm{\nu} = 1 \\
\end{aligned} \end{equation}

Indeed, from Slater's condition, we know that the optimal values of a LP and its dual are equal. Thus the optimal value of LP~\ref{eq:dual} is indeed the optimal $b_0$.

In conclusion, LP~\ref{eq:dual} provides a method to assert rapidly quasi static balance, and returns a robustness value, that can be used as a heuristic for robust static balance.
Another option to enforce robustness is to set a minimum acceptable value for $b_0$; this avoids ponderating balance robustness with other heuristics,
but as a result this conservative approach is not complete.

>>>>

\subsection{A heuristic for robust static equilibrium}
The planner is designed so that any contact configuration it outputs is in static equilibrium.
Equilibrium is of course critical in multipedal locomotion. For this reason we are interested in a robust 
criterion, that ensures that the robot remains balanced in a real world application, regardless of the errors
that can result from perception and control uncertainties.

We first give a linear program (LP) that asserts that a contact configuration is in static equilibrium.
From this formulation we derive a new LP that quantifies the robustness of the configuration regarding equilibrium.
From this value we can in turn either define a heuristic to choose the most robust candidate, or set a treshold on the required robustness
for any configuration. While the LP we propose is original, it is based on an analysis of the problem we proposed in \cite{Prete2016}, where the interested reader
can find more details.

@@

According to Coulomb's law, the non slipping condition is verified if all the contact forces lie in the friction cone defined by the surface.
As classically done, we linearize the friction cone in a conservative fashion with a pyramid, described by four generating rays. We choose for instance:
\begin{equation*}
\mathbf{V}_{i} = \mat{\mathbf{n}_{i} + \mu \mathbf{t}_{i1} & \mathbf{n}_{i} -\mu \mathbf{t}_{i1} & \mathbf{n}_{i} + \mu \mathbf{t}_{i2} & \mathbf{n}_{i} - \mu \mathbf{t}_{i2}}^T
\end{equation*}

Any force belonging to the linearized cone
can thus be express as a convex combination of its four generating rays.
Therefore we can express the non slipping constraint on $\mathbf{f}_i$ as:

\begin{equation*}
\forall i  \qquad  \exists \bm{\beta}_i \in \mathbb{R}^{4} , \bm{\beta}_i \ge 0 \wedge \mathbf{f}_{i} = \mathbf{V}_{i} \bm{\beta}_i \qquad 
\end{equation*}

we can stack the constraints to obtain:
\begin{equation}\label{eq:gen}
\exists \bm{\beta} \in \mathbb{R}^{4e} ,  \bm{\beta} \ge 0 \wedge \mathbf{f} = \mathbf{V} \bm{\beta} \qquad 
\end{equation}
where $\mathbf{V} = \diag{ \{\mathbf{V}_1, \dots, \mathbf{V}_e\} }$, and $\mathbf{f} = [\mathbf{f}_0,...,\mathbf{f}_e]^T$.

From the Newton-Euler equations, and assuming a null acceleration, we also constrain the sum of contact forces to compensate the gravity at the center of mass:

%~ \begin{align}
%~ \underbrace{\mat{m (\ddot{\mathbf{c}} - \mathbf{g})  \\ m \mathbf{c} \times (\ddot{\mathbf{c}} - \mathbf{g}) + \dot{\mathbf{L}}}}_\mathbf{w}
%~ = 
%~ \underbrace{
%~ \mat{\mathbf{I}_3 & \dots & \mathbf{I}_3 \\
%~ \hat{\mathbf{p}}_1 & \dots & \hat{\mathbf{p}}_e} \mathbf{V}
%~ }_\mathbf{G}
%~ \bm{\beta},
%~ \end{align}
%~ where $\mathbf{w}\in \Rv{6}$ is the so-called gravito-inertial wrench (GIW) \citep{qiu:dhm:2011, Caron2015} and $\hat{\mathbf{p}} \in \R{3}{3}$ is the cross-product matrix associated to $\mathbf{p}$.

%~ Static balance assumes zero acceleration thus we can write:
\begin{align} \label{eq:new_eul}
\underbrace{
\mat{\mathbf{I}_3 & \dots & \mathbf{I}_3 \\
\hat{\mathbf{p}}_1 & \dots & \hat{\mathbf{p}}_e} \mathbf{V}
}_\mathbf{G} \bm{\beta}, = 
\underbrace{\mat{\mathbf{0}_{3\times 3} \\ m \hat{\mathbf{g}}}}_{\mathbf{D}} \mathbf{c} + 
\underbrace{\mat{-m\mathbf{g} \\ \mathbf{0}}}_{\mathbf{d}}
\end{align}
where $\hat{\mathbf{x}} \in \R{3}{3}$ is the cross-product matrix associated to $\mathbf{x}$.
The right side of \ref{eq:new_eul} is independent from $\mathbf{c}^z$ because of the cross product matrix $\hat{\mathbf{g}}$ inside $\mathbf{D}$, so we can rewrite it as:
\begin{align} \label{eq:new_eul2d}
\mathbf{w}_0 = \mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d}
\end{align}

@@

\begin{equation} \label{eq:lin_prog} \begin{aligned}
\find \quad & \bm{\beta} \in \Rv{4e} \\
\st \quad &\mathbf{G} \bm{\beta} = \mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d} \\
& \bm{\beta} \ge 0 \\
\end{aligned} \end{equation}

\subsubsection{Formulation of a robust LP:}
Let $b_0 \in \mathbb{R}$ be a scalar value. We now define the following LP:

\begin{equation} \label{eq:lin_prog_rob} \begin{aligned}
\find \quad & \bm{\beta} \in \Rv{4e}, b_0 \in \Rv{} \\
\minimize  \quad & -b_0 \\
\st \quad &\mathbf{G} \bm{\beta} = \mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d} \\
& \bm{\beta} \ge b_0 \\
\end{aligned} \end{equation}

We observe that if $b_0$ is positive, LP~\ref{eq:lin_prog} admits a solution, and $b_0$ is proportional to the minimum distance of the equilibrium contact forces to the boundaries of the friction cones.
If $b_0$ is negative, the configuration is not in static equilibrium, and the value $b_0$ indicates "how far" from equilibrium
the configuration is. $b_0$ can thus be seen as a measure of robustness.

In our implementation, we solve the dual of LP~\ref{eq:lin_prog_rob}.
after a trivial elimination of the Lagrange multipliers associated to the inequality constraints, the dual problem has a lower dimensionality:
\begin{equation} \label{eq:dual} \begin{aligned}
\find \quad & \bm{\nu} \in \Rv{6}\\
\maximize  \quad & -(\mathbf{D}^{xy} \mathbf{c}^{xy} + \mathbf{d})^T \nu \\
\st \quad &\mathbf{G}^T \bm{\nu} \ge 0 \\
& \mathbf{1}^T \mathbf{G}^T \bm{\nu} = 1 \\
\end{aligned} \end{equation}

Indeed, from Slater's condition \citep{Boyd:2004:CO:993483}, we know that the optimal values of an LP and its dual are equal. Thus the optimal value of LP~\ref{eq:dual} is indeed the optimal $b_0$.

In conclusion, LP~\ref{eq:dual} provides a method to assert rapidly static equilibrium, and returns a robustness value.
This robustness criterion can be immediately transformed into a heuristic, and eventually be ponderated with others.
Another option to enforce robustness is to set a minimum acceptable value for $b_0$, but as a result this conservative approach is not complete.

### 2015-12-04 body/intro.tex
corrections andrea

We consider the problem of planning the acyclic sequence of contacts describing the motion of a multiped robot in a cluttered environment. Acyclic contact planning is a particular class of motion planning where every configuration of the resulting trajectory must be in contact with the environment in order to support the balance of the system.

@@

The issue of planning acyclic contacts was first completely described by \citeauthor{conf/iser/BretlRLKA04} in their seminal paper. The issue requires the simultaneous handling of two problems, $\mathcal{P}_1$: a relevant guide trajectory for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: the planning of a discrete sequence of acyclic, balanced contact configurations along the trajectory (A third non trivial problem, $\mathcal{P}_3$, not adressed in this work, then consists in interpolating a complete motion between two postures of the contact sequence).  A key issue is to avoid combinatorial explosion when considering at the same time the possible contact configurations and the potential trajectories. This seminal paper proposes a first effective algorithm, able to handle simple situations (such as climbing scenarios), but not scalable to arbitrary environments. Following it, seve\-ral papers have applied this approach in particular situations, typically limiting the combinatorial by imposing a fixed set of possible contacts \citep{Hauser06usingmotion, stilman2010}.

@@

\textbf{On the other hand, the two problems \Pa and \Pb might be decoupled} to reduce the complexity. The feasibility and interest of the decoupling is shown by \citeauthor{DBLP:conf/iser/EscandeKMG08} who manually set up a rough guide trajectory (i.e. an ad-hoc solution to $\mathcal{P}_1$). \Pb  is then addressed as the combinatorial computation of a feasible contact sequence in the neighborhood of the guide. A solution can then be found, at the cost of prohibitive computation times (several hours). Furthermore, this approach is suboptimal because the quality of the motion is conditioned by the relevance of the guide trajectory,  which is not evaluated \textit{a priori}. \citeauthor{Bouyarmane2009} precisely focus on automatically computing a guide trajectory with guarantees of contact feasibility, by extending key frames of the trajectory into whole-body, balanced contact configurations. Randomly sampled configurations are projected into the contact submanifold using a generalized inverse kinematics solver, a computationally expensive process (about 15 minutes are required to compute a guide trajectory in the examples presented). Moreover this explicit projection is yet an insufficient condition and does not provide strong guarantees on the feasibility of the path between two key positions in the trajectory.

@@

To address \Pb, we consider a truly feasible path as an input, and generate new contact configurations along it using a sampling approach.
The true feasibility of the path provides guarantees that the contacts can be created, and reduce the dimensionality of the problem (at each step the root position is fixed). The theorical completeness of the approach is partially traded for efficiency: instead of sampling online
limb configurations and projecting them onto contact, we precompute offline a large database of limb configurations, stored in a spatial structure optimized
for proximity queries. At runtime, this framework allows to obtain extremly rapidly a set of candidate configurations for contact.
These candidates are sorted using a criterion to evaluate the quasi static balance of the configuration, and ranked according to user defined heuristics.
In our implementation we use two such heuristics, used in all our experiments: a robustness criterion regarding the balance of a configuration, and a manipulability based heuristic,
the Extended FORce Transmission ratio (EFORT).

@@

The present paper is an extension of a conference paper to appear in the proceedings of the ISRR '15 conference TODO CITE.
The conference paper focuses on the theoritical formulation of the problem, and presents results obtained with virtual avatars.
This extension completes this work with an argumented discussion on the implementation of this theory to real world robots and problems.
In particular, we introduce a robust balance criterion, designed to ensure the equilibrium of the robot. We also provide details on the hpp implementation
of our contact generation algorithm. Furthermore, our experiments are now applied to real robot models, namely HRP-2 and Hyq. TODO Appendix describes the generation process
for the robot.

>>>>

We consider the problem of planning the acyclic sequence of contacts describing the motion of a multiped robot in a cluttered environment. Acyclic contact planning is a particular class of motion planning where every configuration of the resulting trajectory must be in contact with the environment in order to support the equilibrium of the system.

@@

The issue of planning acyclic contacts was first completely described by \citeauthor{conf/iser/BretlRLKA04} in their seminal paper. The issue requires the simultaneous handling of two problems, $\mathcal{P}_1$: a relevant guide trajectory for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: the planning of a discrete sequence of acyclic contact configurations along the trajectory (A third non trivial problem, $\mathcal{P}_3$, not adressed in this work, then consists in interpolating a complete motion between two postures of the contact sequence).  A key issue is to avoid combinatorial explosion when considering at the same time the possible contact configurations and the potential trajectories. This seminal paper proposes a first effective algorithm, able to handle simple situations (such as climbing scenarios), but not scalable to arbitrary environments. Following it, seve\-ral papers have applied this approach in particular situations, typically limiting the combinatorial by imposing a fixed set of possible contacts \citep{Hauser06usingmotion, stilman2010}.

@@

\textbf{On the other hand, the two problems \Pa and \Pb might be decoupled} to reduce the complexity. The feasibility and interest of the decoupling is shown by \citeauthor{DBLP:conf/iser/EscandeKMG08} who manually set up a rough guide trajectory (i.e. an ad-hoc solution to $\mathcal{P}_1$). \Pb  is then addressed as the combinatorial computation of a feasible contact sequence in the neighborhood of the guide. A solution can then be found, at the cost of prohibitive computation times (several hours). Furthermore, this approach is suboptimal because the quality of the motion is conditioned by the relevance of the guide trajectory,  which is not evaluated \textit{a priori}. \citeauthor{Bouyarmane2009} precisely focus on automatically computing a guide trajectory with guarantees of contact feasibility, by extending key frames of the trajectory into whole-body contact configurations in static equilibrium. Randomly sampled configurations are projected into the contact submanifold using a generalized inverse kinematics solver, a computationally expensive process (about 15 minutes are required to compute a guide trajectory in the examples presented). Moreover this explicit projection is yet an insufficient condition and does not provide strong guarantees on the feasibility of the path between two key positions in the trajectory.

@@

To address \Pb, we consider a truly feasible path as an input, and generate new contact configurations along it using a sampling approach.
The true feasibility of the path provides guarantees that the contacts can be created, and reduce the dimensionality of the problem (at each step the root position is fixed). The theorical completeness of the approach is partially traded for efficiency: instead of sampling online
limb configurations and projecting them onto contact, we precompute offline a large database of limb configurations, stored in a spatial structure optimized
for proximity queries. At runtime, this framework allows to obtain extremly rapidly a set of candidate configurations for contact.
These candidates are sorted using a criterion to evaluate the quasi static equilibrium of the configuration, and ranked according to user defined heuristics.
In our implementation we use two such heuristics, used in all our experiments: a robustness criterion regarding the static equilibrium of a configuration, and a manipulability based heuristic,
the Extended FORce Transmission ratio (EFORT).

@@

The present paper is an extension of a conference paper to appear in the proceedings of the ISRR'15 conference~\citep(tonneauisrr15).
The conference paper focuses on the theoritical formulation of the problem, and presents results obtained with virtual avatars.
This extension completes this work with an argumented discussion on the implementation of this theory to real world robots and problems.
In particular, we introduce a robust balance criterion, designed to ensure the equilibrium of the robot. We also provide details on the hpp implementation
of our contact generation algorithm. Furthermore, our experiments are now applied to real robot models, namely HRP-2 and Hyq. TODO Appendix describes the generation process
for the robot.
